# Clustering and classification

in this chapter...
please also see [bonus](#bonus) and [shiny](#shiny)

```{r}
library(MASS); library(tidyr); library(dplyr); library(ggplot2); library(corrplot)

# load the data
data("Boston")

# explore the dataset
str(Boston)
summary(Boston)
```

```{r}
# plot matrix of the variables
pairs(Boston)

# calculate the correlation matrix and round it
cor_matrix<-cor(Boston) 

# print the correlation matrix
cor_matrix <- cor_matrix %>% round(2)

# visualize the correlation matrix
corrplot(cor_matrix, method="circle", type="upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6)
```

```{r}
# center and standardize variables
boston_scaled <- scale(Boston)

# summaries of the scaled variables
summary(boston_scaled)

# class of the boston_scaled object
class(boston_scaled)

# change the object to data frame
boston_scaled <- as.data.frame(boston_scaled)

# summary of the scaled crime rate
summary(boston_scaled$crim)

# create a quantile vector of crim and print it
bins <- quantile(boston_scaled$crim)
bins

# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, labels = c("low", "med_low", "med_high", "high"))

# look at the table of the new factor crime
table(crime)

# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)

```


Dividing the data set to train and test sets
```{r}
# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)
# create test set 
test <- boston_scaled[-ind,]

# save the correct classes from test data
correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)
```


## Linear discriminant analysis

More about LDA in this StatQuest-video  
<iframe width="640" height="385" src="https://www.youtube.com/embed/azXCzI57Yfc" frameborder="0" allowfullscreen></iframe>

5. Fit the linear discriminant analysis on the train set. Use the categorical crime rate as the target variable and all the other variables in the dataset as predictor variables. Draw the LDA (bi)plot.
```{r}
# code here
```

```{r}
# code here
```

6. Save the crime categories from the test set and then remove the categorical crime variable from the test dataset. Then predict the classes with the LDA model on the test data. Cross tabulate the results with the crime categories from the test set. Comment on the results
```{r}
# code here
```

```{r}
# code here
```

7. Reload the Boston dataset and standardize the dataset (we did not do this in the Datacamp exercises, but you should scale the variables to get comparable distances). Calculate the distances between the observations. Run k-means algorithm on the dataset. Investigate what is the optimal number of clusters and run the algorithm again. Visualize the clusters (for example with the pairs() or ggpairs() functions, where the clusters are separated with colors) and interpret the results.
```{r}
# code here
```

```{r}
# code here
```

## BONUS{#bonus}

```{r}
# code here
```

```{r}
# code here
```


```{r}
# code here
```

```{r}
# code here
```

# Shiny{#shiny}
```{r}
# code here
```
